National Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs

### 2020-11-29 Manuscript_Bayesian.tex
added saturated models for the bayesian estimate. Not distinguishable from the REML models.
near to final draft, latex table not right yet

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 4685 words in text body, 1191 words in reference section}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi

@@

\abstract{
Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
}

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002). These items are used both to measure spirituality and subjective well-being. As Koenig observes, it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs. Koenig aptly describes this problematic approach as \enquote{tautological} (Koenig, 2008). How pervasive is this problem? A recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016). The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001). Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings. Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012). As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear. Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013). Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of constructs in populations where these dimensions overlap, partially overlap, and diverge.

A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans. Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assessing how their lives turn out (Garssen \& Visser, 2016). Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012). However, To address the key questions in spirituality psychological health research requires longitudinal investigations. Unfortunately the relevant datasets are rare.

Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or life force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force. Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people. Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses).

\hypertarget{measures}{%
\subsection{Measures}\label{measures}}

@@

@@

Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985). Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach's alpha for this 2-item scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017). Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

\hypertarget{belief-in-a-spirit-or-a-god}{%
\subsubsection{Belief in a Spirit or a God}\label{belief-in-a-spirit-or-a-god}}

We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force. These believe were developed for the NZAVS from the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

Table 1 about here:

@@

Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were \enquote{Male} was coded as \enquote{1} and \enquote{Female} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

@@

@@

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses). To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and estimated a random effect for the slope for years. The model equations are given below and priors are given below.

\hypertarget{potential-confounders-and-missing-data}{%
\subsection{Potential Confounders and Missing Data}\label{potential-confounders-and-missing-data}}

Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N \textgreater{} 3 to N = 8 waves. In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: \href{https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf}{NZAVS modelling standards}. We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. There are increasing worries

We provide the analyses of these additional analyses in the supplement. The supplement describing these analysis is located at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.

@@

The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86\_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
We are greatful to the contributors and mantainers of the following packages:
R (Version 4.0.2; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 0.15.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.23; Bates, Mächler, Bolker, \& Walker, 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.8.6.1; D. Lüdecke et al., 2020a), \emph{patchwork} (Version 1.0.1; Pedersen, 2020), \emph{prettycode} (Version 1.1.0; Csárdi, 2019), \emph{purrr} (Version 0.3.4; Henry \& Wickham, 2020), \emph{Rcpp} (Version 1.0.5.1; Eddelbuettel \& François, 2011; Eddelbuettel \& Balamuta, 2017), \emph{readr} (Version 1.3.1; Wickham, Hester, \& Francois, 2018), \emph{report} (Version 0.1.0; Makowski, Dominique, Lüdecke, \& Daniel, 2019), \emph{see} (Version 0.5.2.1; D. Lüdecke et al., 2020b), \emph{sjPlot} (Version 2.8.4; D. Lüdecke, 2020), \emph{stringr} (Version 1.4.0; Wickham, 2019), \emph{styler} (Version 1.3.2; Müller \& Walthert, 2020), \emph{table1} (Version 1.2; Rich, 2020), \emph{texreg} (Version 1.37.5; Leifeld, 2013), \emph{tibble} (Version 3.0.4; Müller \& Wickham, 2020), \emph{tidyr} (Version 1.1.2; Wickham, 2020b), and \emph{tidyverse} (Version 1.3.0; Wickham, Averick, et al., 2019)"

@@

All code for this analysis and a full report of the missing data imputation can be found at \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.

@@

\begin{table}
\begin{center}
\scalebox{0.8}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Sat \\
\midrule
Intercept                           & $\mathbf{7.04}$   & $\mathbf{5.19}$  \\
                                    & $ [ 7.02;  7.07]$ & $ [ 5.17; 5.21]$ \\
yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.02;  0.03]$ & $ [ 0.01; 0.02]$ \\
Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.02$           \\
                                    & $ [-0.09; -0.03]$ & $ [-0.00; 0.04]$ \\
BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.05}$  \\
                                    & $ [-0.10; -0.03]$ & $ [ 0.03; 0.07]$ \\
BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.08}$  \\
                                    & $ [-0.04;  0.04]$ & $ [ 0.04; 0.11]$ \\
yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.00;  0.02]$ & $ [ 0.01; 0.02]$ \\
yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00; 0.02]$ \\
yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$          \\
                                    & $ [-0.01;  0.02]$ & $ [-0.01; 0.01]$ \\
\midrule
SD: Idsd(Intercept)                 & $1.44$            & $0.99$           \\
SD: Idsd(yearsC)                    & $0.13$            & $0.10$           \\
SD: Idcor(Intercept,yearsC)         & $-0.02$           & $0.03$           \\
\bottomrule
\end{tabular}
}
\caption{}
\label{tab:REGRESS}
\end{center}
\end{table}

@@

\begin{figure}
\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefficientplotsbayes} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
\end{figure}

>>>>

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5181 words in text body, 1191 words in reference section}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi

@@

\abstract{
Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or Life Force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or Life Force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
}

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items to measure spirituality such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} (Underwood \& Teresi, 2002). However, these items are also used to measure subjective well-being. Koenig describes this approach as \enquote{tautological} (Koenig, 2008): it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs. How pervasive is the use of tautological measures? A recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016). The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings that some of which apply to non-religious contexts (Swinton, 2001). Notably, spirituality and health research has not converged on a unique set of meanings. Although it is commonplace across most areas of psychological science for researchers to investigate different operationalisations of focal concepts (think of memory research, or emotions research), the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012). Whether the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear. Put simply, cnceptual overlap between spirituality constructs and religion constructs creates a second line of terminological confusion that interacts with terminological confusion arising from overlap between spirituality constructs and well-being constructs. Indeed, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013). Thus to infer how spirituality affects subjective well-being, it is not sufficient to disentange spirituality constructs from tautological well-being constructs, it is also important to disentangle the concept of spirituality from the concept of religion, and to carefully investigate the mechanisms of well-being in populations who overlap, partially overlap, and differ in these dimensions.

A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans. To assess the mechanisms of well-being production requires observing change within people who differ in theoregically relevant ways and assessing how their lives turn out (Garssen \& Visser, 2016). However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have nearly exclusively relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012). Although researching the key questions in spirituality psychological health research requires national longitudinal investigations, unfortunately the relevant datasets are rare.

Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or Life Force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life Force; (4) Those who who believe in a both God and a Spirit or Life Force. Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people. Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses). Descriptive summaries for demographic and theoretical indicators are given in Table 1.

@@

\hypertarget{theoretical-indicators}{%
\subsection{Theoretical indicators}\label{theoretical-indicators}}

@@

Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985). Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach's alpha for this 2-item scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items.

@@

Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were \enquote{Male} was coded as \enquote{1} and \enquote{Female} was coded as \enquote{0.}

@@

Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017). Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items.

\hypertarget{belief-in-a-spiritlife-force-andor-a-god}{%
\subsubsection{Belief in a Spirit/Life Force and/or a God}\label{belief-in-a-spiritlife-force-andor-a-god}}

We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or Life Force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life Force; (4) Believers in a God and Spirit or Life Force. These believe were developed for the NZAVS from the 2005 Eurobarometer.

@@

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (\emph{N} = 20,979 with complete responses). To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years. This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years. We estimated all statitiscal models using both REML estimation (Bates, Mächler, Bolker, \& Walker, 2015) and Bayesian estimation (Bürkner, 2018) with weakly informative priors. We include the full reports in the Supplement. The REML estimates and the Bayesian estimates converged to within three decimal places. We include model equations and priors at the end of this article. Here we report the results of the Bayesian estimator.

\hypertarget{potential-confounders}{%
\subsection{Potential Confounders}\label{potential-confounders}}

To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: \href{https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf}{NZAVS modelling standards}. Next we modelled potential relationships between the coefficients within the model using the daggity package in R, and plotting results using the daggity and ggdag packages in R.

This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g.~retirement). Results were insensitive to the choice set. Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details). In the fully saturated model, it appeared that those skeptical of a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years. Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking. We discuss this point in more detail in the supplement, and mention it here to caution researchers from \enquote{everything-but-the-kitchen-sink} approaches to regression modelling.

\hypertarget{missing-data}{%
\subsection{Missing Data}\label{missing-data}}

Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N \textgreater{} 3 to N = 8 waves.
We include the multiple-imputation results in a supplement. Here, there were no differences in estimation relevant to inference, however, to improve inference, we advocate wider use of multiple-imputation or Bayesian approaches for handling missingness.

@@

The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86\_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
We are greatful to the contributors and mantainers of the following packages:
R (Version 4.0.2; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 0.15.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.23; Bates et al., 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.8.6.1; D. Lüdecke et al., 2020a), \emph{patchwork} (Version 1.0.1; Pedersen, 2020), \emph{prettycode} (Version 1.1.0; Csárdi, 2019), \emph{purrr} (Version 0.3.4; Henry \& Wickham, 2020), \emph{Rcpp} (Version 1.0.5.1; Eddelbuettel \& François, 2011; Eddelbuettel \& Balamuta, 2017), \emph{readr} (Version 1.3.1; Wickham, Hester, \& Francois, 2018), \emph{report} (Version 0.1.0; Makowski, Dominique, Lüdecke, \& Daniel, 2019), \emph{see} (Version 0.5.2.1; D. Lüdecke et al., 2020b), \emph{sjPlot} (Version 2.8.4; D. Lüdecke, 2020), \emph{stringr} (Version 1.4.0; Wickham, 2019), \emph{styler} (Version 1.3.2; Müller \& Walthert, 2020), \emph{table1} (Version 1.2; Rich, 2020), \emph{texreg} (Version 1.37.5; Leifeld, 2013), \emph{tibble} (Version 3.0.4; Müller \& Wickham, 2020), \emph{tidyr} (Version 1.1.2; Wickham, 2020b), and \emph{tidyverse} (Version 1.3.0; Wickham, Averick, et al., 2019)"

@@

All code for this analysis and a full report of the missing data imputation can be found at \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}. An html document decribing the same analysis is located at: \url{https://osf.io/cu2gr/}.

@@

\begin{table}
\begin{center}
\scalebox{0.5}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Sat \\
\midrule
Intercept                           & $\mathbf{7.14}$   & $\mathbf{5.29}$   \\
                                    & $ [ 7.11;  7.18]$ & $ [ 5.26;  5.31]$ \\
yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.01]$ \\
Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.01$            \\
                                    & $ [-0.09; -0.03]$ & $ [-0.01;  0.03]$ \\
BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.03}$   \\
                                    & $ [-0.09; -0.03]$ & $ [ 0.01;  0.06]$ \\
BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.06}$   \\
                                    & $ [-0.04;  0.04]$ & $ [ 0.03;  0.09]$ \\
Age\_in\_Decades\_C                 & $\mathbf{0.07}$   & $\mathbf{0.08}$   \\
                                    & $ [ 0.06;  0.09]$ & $ [ 0.07;  0.09]$ \\
Ethnic\_CategoriesMaori             & $\mathbf{-0.38}$  & $\mathbf{-0.11}$  \\
                                    & $ [-0.43; -0.33]$ & $ [-0.14; -0.08]$ \\
Ethnic\_CategoriesPacific           & $\mathbf{-0.38}$  & $\mathbf{-0.15}$  \\
                                    & $ [-0.48; -0.29]$ & $ [-0.22; -0.09]$ \\
Ethnic\_CategoriesAsian             & $\mathbf{-0.09}$  & $-0.05$           \\
                                    & $ [-0.18; -0.01]$ & $ [-0.11;  0.02]$ \\
Male1                               & $\mathbf{-0.06}$  & $\mathbf{-0.16}$  \\
                                    & $ [-0.10; -0.02]$ & $ [-0.19; -0.13]$ \\
yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
                                    & $ [ 0.00;  0.02]$ & $ [ 0.00;  0.02]$ \\
yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.02]$ \\
yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$           \\
                                    & $ [-0.01;  0.02]$ & $ [-0.01;  0.01]$ \\
\midrule
SD: Idsd(Intercept)                 & $1.42$            & $0.98$            \\
SD: Idsd(yearsC)                    & $0.13$            & $0.10$            \\
SD: Idcor(Intercept,yearsC)         & $-0.03$           & $0.03$            \\
\bottomrule
\end{tabular}
}
\caption{}
\label{tab:REGRESS_LS}
\end{center}
\end{table}

@@

\begin{figure}
\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
\end{figure}
